% declare document class and geometry
\documentclass[12pt]{article} % use larger type; default would be 10pt
\usepackage[margin=1in]{geometry} % handle page geometry

% import packages and commands
\input{../header2.tex}


\title{Math 217 -- Geometry and Physics -- Lec03}
\author{UCLA, Fall 2014}
\date{\formatdate{08}{10}{2014}} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed 

\begin{document}
\maketitle


\section{Stuff}

Seminal paper: Witten --- Supersymmetry and Morse Theory (J. Diff. Geometry, 1982)

\subsection{Chern conjecture}

Let's go over the Chern conjecture from last time. If $M = M^{2n}$ is a compact closed manifold, then
\begin{equation}
\chi(M) = \int_M \varepsilon(TM).
\end{equation}
Given the metric $g$ on $TM$, we have the Levi-Civita connection $\nabla^g$. Then we have $R^{\nabla^g} = (\nabla^g)^2$ where $R$ is skew-symmetric, and 
\begin{eqn}
\abs{Pf(R^{\nabla^g})} = \varepsilon(TM) \in H^{2n}(M).
\end{eqn}
[not sure transcribed correctly, was fast and messy on the board.] If $R^{\nabla^g} = 0$ then $\chi(M)= 0$ where 
\begin{eqn}
\chi(M) = \sum_{i=0}^{2n} (-1)^i b_i, \qquad b_i = \dim_\R H^i (M).
\end{eqn}

The Chern conjecture is based on that. Conjecture: If $M^{2n}$ is an affine flat closed compact manifold, then $\chi(M) = 0$. Let's define what affine flat means. Write $M = \cup_{\alpha \in I} U_\alpha$. Then it is affine flat iff
\begin{eqn}
\varphi_\beta \circ \varphi_\alpha^{-1}(x_\alpha) = A_{\alpha\beta} x_\alpha + B_{\alpha\beta},
\end{eqn}
where $A_{\alpha\beta}, B_{\alpha\beta}$ are constant matrices $\iff$ there exists an affine connection 
\begin{eqn}
\nabla : \Gamma(TM) \rightarrow \Gamma(T^* M \otimes TM)
\end{eqn}
where $\nabla(fs) = f \nabla s + f \nabla s$. 

The Chern conjecture is known in 2 dimensions (the torus). Not sure if known in any other dimensions. (See Sullivan-Kostant-Milnor for special cases.) 


\subsection{Poincare duality}

Says that $H^p \cong H^{n-p}$. 

\begin{exercise}
As an exercise, show (cf. Bott-Tu):
\begin{enumerate}
\item $M = \R^n$. 
\begin{equation}
	H^k (\R^n) = 
	\begin{cases}
	0, & k = 0, \\
	\R, & k \neq 0
	\end{cases}
	\qquad
	H_c^k (\R^n) = 
	\begin{cases}
	0, & k \neq n \\
	\R, & k = n.
	\end{cases}
\end{equation}
\item $M = \cup_{\alpha \in I} U_\alpha$. 
\end{enumerate}
\end{exercise}

Now, given $\Delta \subseteq M$ a $p$-cycle in $M$, we have 
\begin{eqn}
[\Delta] \in H_p (M, \R) \cong H^{n-p}(M, \R)
\end{eqn}
Then we can write 
\begin{eqn}
\begin{matrix}
\int_\Delta \omega & : & H_\text{dR}^p(M) & \rightarrow & \R \\
&& [\omega] & \mapsto & \int_\Delta \omega
\end{matrix}
\end{eqn}
for a $p$-form $\omega$. Write $\Delta = \sum_i a_i \underbrace{\Delta_i}_{=\varphi_i(\tilde{\Delta}_i)}$ where $\tilde{\Delta}_i \subseteq [\text{something}] \overset{\varphi_i}{\longrightarrow} M$


\subsection{stuff}

[missed a lot here on whatever he was talking about]


\subsection{Lie groups}

\begin{definition}
A Lie group $G$ is a smooth manifold with a group structure in the following sense:
\begin{enumerate}
\item Multiplication $G \times G \rightarrow G$ taking $(g, g') \mapsto gg'$, and
\item Inverse $G \rightarrow G$ taking $g \mapsto g^{-1}$
\end{enumerate}
\end{definition}

Some examples are given by: $\C, \C, \Q, \Q_p, \dots$. Write $M_n(K) = \text{$n \times n$ matrices with entries in $K$}$. The group with elements $g = (a_{ij})_{n \times n} \in GL_n(\R)$ with $\det g \neq 0$ is called the general linear group. Note that any subgroup submanifold is also a Lie group. 

\begin{definition}
A Lie subgroup $H$ of $G$ is a subgroup which is also a submanifold. If $H \cong \R$ then $i : \R \rightarrow G$ defines a one-parameter subgroup. 
\end{definition}

Some more examples: $SL_n(\R), O_n(\R), SO_n(\R), U_n, SU_n$

\begin{definition}
A Lie group homomorphism $\psi : G \rightarrow G'$ is a map satisfying $\psi(gg') = \psi(g) \psi(g')$. If $\psi$ is a diffeomorphism then it is called an isomorphism. It will be useful to also define
\begin{itemize}
\item $R_g : G \rightarrow G$ mapping $R_g (g') = g' g$ for all $g' \in G$
\item $L_g : G \rightarrow G$ mapping $L_g (g') = g g'$ for all $g' \in G$.
\end{itemize}
\end{definition}

\begin{definition}
A vector field $X$ on $G$ is said to be left invariant if $L_g X = X$.
\end{definition}








\end{document}
