% declare document class and geometry
\documentclass[12pt]{article} % use larger type; default would be 10pt
\usepackage[margin=1in]{geometry} % handle page geometry

% import packages and commands
\input{../header2.tex}

\newcommand{\Gr}{\opname{Gr}}


\title{Math 217 -- Geometry and Physics -- Lec12}
\author{UCLA, Fall 2014}
\date{\formatdate{29}{10}{2014}} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed 

\begin{document}
\maketitle


\section{More stuff}

Recall Riemann-Roch. Given a function $f : M \rightarrow N = p+$, define a function
\begin{eqn}
f_! : K(M) \rightarrow K(p+) \cong \Z
\end{eqn}
where $K(p+) \ni [E] \cong \rank E = r$. Then we have $\hat{A} (Tp+) = 1$, and
\begin{eqn}
\int_M \hat{A} (TM) \opname{ch} E \in \Z
\end{eqn}
for all $[E] \in K(M)$. If $M$ is a spin manifold, then
\begin{eqn}
\int_M \hat{A} (TM) \in \Z.
\end{eqn}
Recall of course 
\begin{eqn}
\hat{A} (TM) = \frac{ \prod_j^n x_j }{ \prod_j^n (e^{-x_j/2} - e^{-x_j/2}) },
% error? denom = 0.. check previous notes
\end{eqn}
where the numerator is just $e(TM)$ and the denominator is the Chern character of the ``Euler class'' of $TM$ in $K(M)$. 

If $f : M \rightarrow N$ is a holomorphic map between projective manifolds, then (GRR)
\begin{eqn}
f_* (T \dif(TM) \opname{ch} E) = T \dif(TN) \opname{ch} f_! E.
\end{eqn}
We have 
\begin{eqn}
T \dif(TM^n) = \frac{ \prod_j^n x_j}{\prod_j^n (1-e^{-x_j})},
\end{eqn}
where again the numerator is just $e(TM)$ and the denominator is the Chern character of Euler class of $TM$ in $K(M)$. When $N = p+$, 
\begin{eqn}
f_! E = \sum_{q=0}^n (-1)^q \dim H^q (M,E) \in K(p+) \cong \Z.
\end{eqn}
Hirzebruch RR tells us that this sum is just
\begin{eqn}
f_! E = \int_M T \dif(TM) \opname{ch} E.
\end{eqn}


\begin{definition}[Cobordism (Generalized cohomology theory)]
Given oriented, closed, compact manifolds $M_1, M_2$, they are said to be cobordant $M_1 \sim M_2$ if there exists a manifold $M$ with boundary 
\begin{eqn}
\partial M = M_1 \cup -M_2,
\end{eqn}
where $-M_2$ denotes $M_2$ with the reverse orientation. We denote $\Omega_\text{SO}^*$ as the set of cobordism equivalence classes. 
\end{definition}
% why SO? why use U below?

These equivalence classes have a ring structure,
\begin{align}
[M_1] \times [M_2] &= [M_1 \times M_2] \\
[M_1] + [M_2] &= [M_1 \sqcup M_2],
\end{align}
where $\sqcup$ denotes disjoint union. 

\begin{theorem}
We have
\begin{eqn}
\Omega^*_\text{SO} \otimes_\Z \Q = \Omega^*_U \otimes_\Z \Q = \Q[\CP^2, \CP^4, \dots].
\end{eqn}
(Due to Thom Milnor. [?])
\end{theorem}

% $J^2 = -I$ on $M$ then $TM \oplus \R^k$ has `almost' complex structure.  [huh?]

\begin{lemma}
The pontryagin numbers are cobordism invariants.
\end{lemma}

\begin{proof}
Given $\partial M \overset{i}{\hookrightarrow} M$, we have
\begin{eqn}
P(i^* TM) = P(T \partial M) \underbrace{P(\R)}_{=1},
\end{eqn}
so Stoke's theorem tells us
\begin{eqn}
\int_{\partial M} i^* \omega = \int_M \dif{\omega} = 0
\end{eqn}
for any pontryagin class $\omega$.
\end{proof}

A genus is a ring homomorphism:
\begin{eqn}
\varphi : \Omega^*_\Q = \Q[\CP^2, \CP^4, \dots] \rightarrow R \text{ [ring]},
\end{eqn}
where we have
\begin{align}
\varphi([M] \times [N]) &= \varphi([M]) \varphi([N]), \\
\varphi([M] + [N]) &= \varphi([M]) + \varphi([N]).
\end{align}

\begin{example} %is it example?
Let $f(x) = \frac{x}{Q(x)}$ where $Q(x)$ is an even function and so $f(x)$ is odd and we can write
\begin{eqn}
f(x) = b_1 x + b_3 x^3 + \dots, \qquad g(x) = 1 + a_2 x^2 + a_4 x^4 + \dots.
\end{eqn}
Then consider
\begin{eqn}
\varphi([M^{2n}]) \equiv \int_M \prod_{j=1}^n Q(x_j)
\end{eqn}
where $\set{x_j}$ are the Chern roots of $TM$. Then $\varphi$ is a genus associated to $f$.
\end{example}

\begin{lemma}
Let $y = f(x)$, $g = f\inv$. Then
\begin{eqn}
g'(y) = \sum_{n \geq 0}^\infty \varphi(\CP^n) y^n
\end{eqn}
where
\begin{eqn}
\varphi(M) = \int_M \prod_{j=1}^n Q(x_j).
\end{eqn}
\end{lemma}

\begin{proof}
We have
\begin{eqn}
T \CP^n \oplus \C \cong \underbrace{\bigO(1) \oplus \dots \oplus \bigO(1)}_{n+1},
\end{eqn}
and so
\begin{eqn}
c(T \CP^n) = c(\bigO(1))^{n+1} = t^{n+1}.
\end{eqn}
If we denote $t = c(\bigO(1))$, then
\begin{eqn}
H^* (\CP^n, \Z) = \Z[t] \Big/ t^{n+1} = 0
\end{eqn}
[???] and we have $\int_{\CP^n} t^n = 1$ and
\begin{align}
\varphi(\CP^n) &= \int_{\CP^n} Q(t)^{n+1} = \int_{\CP^n} \left( \frac{t}{f(t)} \right)^{n+1} = \Res\limits_{x=0} \left( \frac{1}{f(x)} \right)^{n+1} \\
	&= \frac{1}{2\pi i} \int_C \left( \frac{1}{f(x)} \right)^{n+1} \dif{x} = \frac{1}{2\pi i} \int_{f(C)} \frac{g'(y)}{y^{n+1}} \dif{y} \\
	&= \text{coeff. of $y^n$ in $g'(y)$},
\end{align}
where we have substituted $y = f(x)$, $\dif{y} = f'(x) \dif{x}$, $f'(x) = 1 / g'(y)$. 
\end{proof}

\begin{example}
Take
\begin{eqn}
Q(x) = \frac{x}{\tanh x}, \qquad f(x) = \tanh x.
\end{eqn}
Then 
\begin{eqn}
\frac{1}{g'(y)} = f'(x) = 1 - f(x)^2 \quad \implies \frac{1}{g'(y)} = 1 - y^2 \quad \implies g'(y) = \frac{1}{1-y^2} = 1 + y^2 + y^4 + \dots
\end{eqn}
so we have
\begin{eqn}
L(\CP^n) = 
\begin{cases}
0 & \text{$n$ odd}, \\
1 & \text{$n$ even}.
\end{cases}
\end{eqn}
\end{example}

\begin{theorem}[Thom]

[missed this]

\end{theorem}










\end{document}
