% declare document class and geometry
\documentclass[12pt]{article} % use larger type; default would be 10pt
\usepackage[margin=1in]{geometry} % handle page geometry

% import packages and commands
\input{../header2.tex}

\newcommand{\Gr}{\opname{Gr}}


\title{Math 217 -- Geometry and Physics -- Lec13}
\author{UCLA, Fall 2014}
\date{\formatdate{31}{10}{2014}} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed 

\begin{document}
\maketitle


\section{More stuff}

Hirzebruch theorem, arguably most important in subject:
\begin{eqn}
\Omega^*_\Q = \Q[\CP^2, \CP^4, \dots]
\end{eqn}
and
\begin{eqn}
L(\CP^{2n}) = 1 = \opname{sgn} M.
\end{eqn}
This implies that 
\begin{eqn}
L(M) = \opname{sgn} M
\end{eqn}
which is zero unless $M$ is $4k$-dimensional. Recall that for
\begin{eqn}
Q(x) = \frac{x/2}{\sinh (x/2)},
\end{eqn}
we have
\begin{eqn}
f(x) = 2 \sinh (x/2)
\end{eqn}
and
\begin{eqn}
f'(x)^2 = 1 + \frac{1}{4} f(x)^2, \qquad g'(y) = (1 + \frac{1}{4} y)^{-1/2}.
\end{eqn}
We find that 
\begin{eqn}
\varphi(\CP^n) = \hat{A} (\CP^n) = 
\begin{cases}
0 & \text{$n$ odd} \\
-1/8 & \text{$n$ even}.
\end{cases}
\end{eqn}
Note that $CP^{2k}$ is not a spin manifold. 

A hot topic for the last 20 or so years has been the following
\begin{example}
Suppose $f(x)$ has derivative of the form
\begin{eqn}
f'(x)^2 = 1 - 2 \delta f^2 + \epsilon f^4,
\end{eqn}
i.e. 
\begin{eqn}
y^2 = 1 - 2 \delta x^2 + \epsilon x^4
\end{eqn}
which is an elliptic curve. Then $Q(x) = x / f(x)$ is an elliptic genus. For example if $\delta = 1$, $\epsilon = 0$ [error, $\delta$ does not match up], then we have
\begin{eqn}
f'(x) = 1 - f^2
\end{eqn}
so that $f(x) = \tanh x$. If $\epsilon = 0$, $\delta = -1/8$, then we have
\begin{eqn}
f'(x) = 1 + \frac{1}{4} f^2,
\end{eqn}
so that $f(x) = 2 \sinh (x/2)$. As an exercise, find the elliptic genus $\varphi_\text{ell}(\CP^n)$. 
\end{example}

This implies that we have elliptic cohomology. Given $\Omega_U^*$ we have a universal group loaw (Quiller). Given genus
\begin{eqn}
\varphi : \Omega_U^* (p+) \rightarrow R
\end{eqn}
where $R$ is the coefficient ring of a cohomology. Then
\begin{eqn}
H^* (M, R) \cong \Omega^*_U (M) \oplus_{\Omega^*_U (p+)} R.
\end{eqn}

We can consider analogous geometric constructions
\begin{itemize}
\item Cohomology --- singular cycles, $H^*_\text{dR}$ differential forms.
\item $K$-theory --- Bundles
\item Cobordism manifolds --- [?]
\item ? Elliptic cohomology --- 
\begin{eqn}
Ell^*(M) = \Omega_U^* (M) \oplus_{\omega^*_U} R \qquad \implies \qquad R = \Z [\frac{1}{2}, \epsilon, \delta]
\end{eqn}
where 
\begin{eqn}
\varphi(M^{2k}) = \int_M \prod_{j=1}^n Q(x_j).
\end{eqn}
\end{itemize}


\section{Elliptic operators}

\begin{definition}
Consider two complex vector bundles $E,F \rightarrow M$ on a smooth closed orientable manifold $M$. Given smooth sections $\Gamma(E), \Gamma(F)$ a map
\begin{eqn}
D : \Gamma(E) \rightarrow \Gamma(F)
\end{eqn}
is called a differential operator on $U \subseteq M$ iff $D$ can be written
\begin{eqn}
Df = g.
\end{eqn}
If we write $f = (f_1, \dots, f_m)^\top$ and $g = (g_1, \dots, g_n)^\top$ then we can write $D$ as a matrix
\begin{eqn}
D_{ij} = \sum_I a^I_{ij} \pd[I]{}{x}, \qquad I = (i_1, \dots, i_n).
\end{eqn}
\end{definition}

\begin{definition}
We define the order $p$ of $D$ as the highest degree of the derivatives in $D$, and the symbol $\sigma(D)$ of $D$ is a function
\begin{eqn}
\sigma(D) = \sigma^{(p)} (D) : \pi^* E \rightarrow \pi^* F
\end{eqn}
defined by
\begin{eqn}
\sigma(D) (x,v) e = D \left( \frac{i^p}{p!} ((h - h(x))^p f) \right) (x)
\end{eqn}
where $e = f(x)$ and $f \in \Gamma(E)$. 
\end{definition}

\begin{example}
Consider $M = \R^k / L$ where $L \subset \R^k$ (a lattice of rank $k$). Given 
\begin{eqn}
E = F = \C \times M \qquad \implies \Gamma(E) = \Gamma(F) = C^\infty (M),
\end{eqn}
we have
\begin{eqn}
D = \Delta = \sum_{i=1}^k \pd[2]{}{x_i}.
\end{eqn}
Then given
\begin{eqn}
v = \sum_{i=1}^k v_i \dif{x_i} \in T_x^* M,
\end{eqn}
we have
\begin{eqn}
\sigma(D) = \sigma^{(2)} (x,v) = -\sum_{i=1}^k v_i^2.
\end{eqn}
\end{example}

\begin{definition}
$D$ is called elliptic iff $\sigma(D)$ is an isomorphism for all $v \neq 0$.
\end{definition}

One can show that if $D$ is elliptic then 
\begin{eqn}
\ker D \subseteq \Gamma(E), \qquad \opname{coker} D = \Gamma(F) / \Im D.
\end{eqn}
are finite dimensional spaces.

\begin{definition}
The index $\opname{Ind} D$ of $D$ is defined by
\begin{eqn}
\opname{Ind} D = \dim_\C \ker D - \dim \opname{coker} D.
\end{eqn}
\end{definition}

So if $D$ is elliptic then $\sigma(D)$ defines an element 
\begin{eqn}
\eta_D = (\pi^* E, \pi^* F, \sigma(D)) \in K(T^* M, T^* M - \set{0}).
\end{eqn}

\begin{remark}
\begin{eqn}
K(T^* M, T^* M - \set{0}) \cong K(M).
\end{eqn}
Given 
\begin{eqn}
f : M \rightarrow p+
\end{eqn}
so
\begin{eqn}
\begin{matrix}
f_! & : & K(M) & \rightarrow & K(p+) \cong \Z \\
& & \eta_D & \mapsto & f_! \eta_D.
\end{matrix}
\end{eqn}
Atiyah-Singer showed
\begin{eqn}
\opname{Ind} D = \opname{ch} f_! \eta_D = \int_M (\text{char. classes})
\end{eqn}
the last equality by topological RR. 
\end{remark}





\end{document}
