% declare document class and geometry
\documentclass[12pt]{article} % use larger type; default would be 10pt
\usepackage[margin=1in]{geometry} % handle page geometry

% import packages and commands
\input{../header2.tex}

% title information
\title{Phys 221A -- Quantum Mechanics -- Lec05}
\author{UCLA, Fall 2014}
\date{\formatdate{20}{10}{2014}} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed 

\begin{document}
\maketitle


[missed first hour, lots of stuff missing here]

\section{Spin precession (of a localized spin-$1/2$ particle)}

We have a 2 dimensional Hilbert space with Hamiltonian
\begin{eqn}
H = -\gamma (\hbar / 2) \v{\sigma} \cdot \v{B},
\end{eqn}
where $\gamma$ is the gyromagnetic ratio and
\begin{eqn}
\v{\sigma} = \set{ \pmat{0 & 1 \\ 1 & 0}, \pmat{0 & -i \\ i & 0}, \pmat{1 & 0 \\ 0 & -1} }
\end{eqn}
are the Pauli matrices. We can also write
\begin{eqn}
H = -\gamma \v{S} \cdot \v{B} = -\v{\mu} \cdot \v{B},
\end{eqn}
where $\v{S} = (\hbar/2) \v{\sigma}$ is the spin operator and $\v{\mu} = \gamma \v{S}$ is the magnetic moment. 

For the electron, we have
\begin{eqn}
\mu = \gamma \hbar / 2 = -(g/2) \mu_B
\end{eqn}
where $g$ is the so-called ``g factor'' and $\mu_B = \abs{e} \hbar / (2mc)$ is the Bohr magneton. Classically it is expected that $g = 1$, which was actually measured (incorrectly) by Einstein et al. According to QED and the Dirac equation, $g \approx 2$ for a free electron. In fact it is very close to 2, actually around $g \approx 2.002$ which has been measured with great accuracy in agreement with theory. This measurement is one of the great triumphs of quantum field theory. 

\subsubsection{Constant magnetic field}

Let $\v{B} = B \uv{z}$, then
\begin{eqn}
H = -\gamma (\hbar / 2) B \v{\sigma}_z = \pmat{-\mu B & 0 \\ 0 & +\mu B}.
\end{eqn}
So we have two energy eigenkets,
\begin{eqn}
\ket{\uparrow} = \pmat{1 \\ 0}, \qquad \ket{\downarrow} = \pmat{0 \\ 1}
\end{eqn}
with energies $E_\uparrow = -\mu B$, $E_\downarrow = \mu B$. Since $\mu < 0$, $E_\downarrow$ is the energy of the ground state. Furthermore, we have
\begin{eqn}
\Delta E = E_\uparrow - E_\downarrow = -2\mu B = \hbar \abs{\gamma} B
\end{eqn}
or $\Delta E = \hbar \omega_L$ where $\omega_L = \abs{\gamma} B$ is the Larmor frequency. 

Now, for an arbitrary state $\ket{\psi}$ we can write
\begin{eqn}
\ket{\psi} = C_\uparrow \ket{\uparrow} + c_\downarrow \ket{\downarrow},
\end{eqn}
which time evolved becomes
\begin{align}
\ket{\psi(t)} &= e^{-(i / \hbar) E_\uparrow t} c_\uparrow \ket{\uparrow} + e^{-(i / hbar) E_\downarrow t} c_\downarrow \ket{\downarrow} \\
	&= e^{-(i / \hbar) E_\uparrow t} (c_\uparrow \ket{\uparrow} + e^{i\omega_L t} c_\downarrow \ket{\downarrow}) \\
	&\sim c_\uparrow \ket{\uparrow} + e^{i\omega_L t} c_\downarrow \ket{\downarrow}.
\end{align}
The expectation value of spin angular momentum is given by
\begin{eqn}
\v{s}(t) = \frac{\hbar}{2} \bra{\psi(t)} \v{\sigma} \ket{\psi(t)},
\end{eqn}
which satisfies the equation
\begin{eqn}
\od{\v{s}(t)}{t} = -\gamma \v{B} \times \v{s}(t) = \omega_L \uv{z} \times \v{s}(t).
\end{eqn}


\section{Heisenberg picture}

In the Schroedinger picture, the operators are time-independent while the state evolves in time. In the Heisenberg picture, we picture the situation the other way around, that is the states are time-independent and the operators evolve in time. In general, for an operator $O(t)$ the time average is given by
\begin{eqn}
\avg{O(t)} = \bra{\psi(t)} O(t) \ket{\psi(t)}.
\end{eqn}
In the Schroedinger picture, we solve for $\ket{\psi(t)}$. In the Heisenberg picture, we have
\begin{align}
\avg{O(t)} &= \bra{\psi} U^\dagger(t) O(t) U(t) \ket{\psi} \\
	&= \bra{\psi} O_H (t) \ket{\psi},
\end{align}
where 
\begin{eqn}
O_H(t) = U^\dagger (t) O(t) U(t)
\end{eqn}
is the operator in the Heisenberg picture. 




\end{document}
