% declare document class and geometry
\documentclass[12pt]{article} % use larger type; default would be 10pt
\usepackage[margin=1in]{geometry} % handle page geometry

% import packages and commands
\input{../header.tex}

% title information
\title{Phys 221A -- Quantum Mechanics -- Lec04}
\author{UCLA, Fall 2014}
\date{\formatdate{15}{10}{2014}} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed 

\begin{document}
\maketitle


\section{Continuous variables}

It's possible to choose an orthonormal basis $\set{\ket{\xi}}$ such that 
\begin{equation}
\braket{\xi}{\xi'} = \delta(\xi - \xi').
\end{equation}
Then for any $\ket{\psi} \in G$ we have
\begin{equation}
\ket{\psi} = \int d\xi ~ c_\xi \ket{\xi}.
\end{equation}
This immediately implies the closure relation
\begin{equation}
\int d\xi \ket{\xi} \bra{\xi} = 1,
\end{equation}
and furthermore we have
\begin{equation}
\braket{\xi'}{\psi} = \int d\xi c_\xi \braket{\xi'}{\xi} = c_{\xi'}.
\end{equation}
We can check the closure relation:
\begin{equation}
\int d\xi \ket{\xi} \braket{\xi}{\psi} = \int d\xi ~ c_\xi \ket{\xi} = \ket{\psi},
\end{equation}
which is what we wanted.

Now, given an operator $O$, we can write $O(\xi, \xi') = \bra{\xi} O \ket{\xi'}$. If we have $O\ket{\psi} = \ket{\psi'}$ and 
\begin{equation}
\int d\xi ~ c_\xi \ket{\xi}, \qquad \int d\xi ~ b_\xi \ket{\xi},
\end{equation}
then we have
\begin{eqn}
b_\xi = \int d\xi' ~ O(\xi, \xi') c_{\xi'}.
\end{eqn}

In the position representation we have $\xi \rightarrow \v{r}$ in $n$-dimensional space. So
\begin{eqn}
\ket{\psi} \rightarrow c_{\v{r}} = \braket{\v{r}}{\psi} \equiv \psi(\v{r})
\end{eqn}
or we can write
\begin{eqn}
\ket{\psi} = \int d^3r ~ \psi(\v{r}) \ket{\v{r}}.
\end{eqn}
In the momentum representation we would have
\begin{eqn}
\ket{\psi} = \int d^3p ~ \psi(\v{p}) \ket{\v{p}}.
\end{eqn}
Then if we measure the position, the probability density is given by $\abs{\psi(\v{r})}^2$ and the probability to find the particle in a given volume $dV$ is given by $\abs{\psi(\v{r})}^2 dV$. This of course assumes normalization,
\begin{eqn}
\int d^3r \abs{\psi(\v{r})}^2 = 1, \qquad \text{or} \qquad \braket{\psi}{\psi} = 1.
\end{eqn}


\section{Spinless particle in 1D space}

Classically we would have the Lagrangian $\mathcal{L}(x, \dot{x}, t)$. Quantum mechanically, we will have a Hilbert space $L_2$ over $c$-valued function $f(x)$. That is, functions in the space have the property
\begin{eqn}
\int dx \abs{f(x)}^2 < \infty,
\end{eqn}
with inner product given by
\begin{eqn}
\braket{f}{g} = \int dx f^*(x) g(x).
\end{eqn}
Physical states will be normalized in this space,
\begin{eqn}
\braket{\psi}{\psi} = 1, \qquad \text{or} \qquad \int dx \abs{\psi(x)}^2 = 1.
\end{eqn}

Let's look at some operators in this space. We find that $x$ is a Hermitian operator:
\begin{eqn}
\int dx (x f(x))^* g(x) \int dx f^*(x) (x g(x)).
\end{eqn}
We find that $d/dx$ is anti-Hermitian:
\begin{eqn}
\int dx (\frac{d}{dx} f(x))^* g(x) = \int dx f^*(x) (-\frac{d}{dx}) g(x)
\end{eqn}
so that $(d/dx)^\dagger = -d/dx$ and $i d/dx$ is Hermitian. 

We will, for now, \textit{define} the linear momentum: $p = -i \hbar d/dx$. Later we will see that this reduces to the classical momentum. Furthermore, the momentum part of the Hamiltonian will typically be given by
\begin{eqn}
H = \frac{p^2}{2m} = -\frac{\hbar}{2m} \partial_x^2.
\end{eqn}
In general $H(\v{r},\v{p})$ will be the same in classical and quantum mechanics. 

We also have the commutation relation
\begin{eqn}
[x,p] = i\hbar. 
\end{eqn}
This gives us the Heisenberg uncertainy principle,
\begin{eqn}
\avg{(\Delta x)^2} \avg{(\Delta p)^2} \geq \hbar^2 / 4.
\end{eqn}
A Gaussian wave packet will take the form
\begin{eqn}
\psi(x) = \frac{1}{\pi^{1/4} \sqrt{d}} ~ e^{ikx - x^2 / 2d}
\end{eqn}
for all $k, d \in \reals$ such that $d > 0$. Gaussian wave packets minimize the uncertainty,
\begin{eqn}
\avg{(\Delta x)^2} \avg{(\Delta p)^2} = \hbar^2 / 4.
\end{eqn}

In the position representation we might have an equation
\begin{eqn}
x g_y(x) = y g_y(x) 
\end{eqn}
for any fixed $y \in \reals$. Then $y$ is the eigenvalue of $x$ and $g_y(x)$ is the eigenfunction of $x$, and we have
\begin{eqn}
g_y(x) = \delta(x-y) \equiv \ket{y}
\end{eqn}
and 
\begin{eqn}
\braket{y}{y'} = \int dx \delta(x-y) \delta(x-y') = \delta(y-y').
\end{eqn}

Next we have the eigenvalue equation
\begin{eqn}
(-i \hbar \frac{d}{dx}) \ket{f_p} = p \ket{f_p}.
\end{eqn}
The eigenstate will be given by 
\begin{eqn}
f_p (x) = e^{ipx / \hbar} / \sqrt{2\pi \hbar}
\end{eqn}
where we can write a wavelength $\lambda = 2 \pi \hbar / p$ which is just the de Broglie wavelength. Furthermore we have
\begin{eqn}
\braket{f_p}{f_{p'}} = \int dx \frac{1}{2\pi \hbar} e^{i(p-p') x / \hbar} = \delta(p - p').
\end{eqn}
Recall that we can write
\begin{eqn}
\delta(k) = \int \frac{dx}{2\pi} e^{ikx}.
\end{eqn}

We can write down a translation operator in $L_2$,
\begin{eqn}
U(a) = e^{-a \partial_x} = 1 - a \partial_x + (a/2) \partial_x^2 + \dots
\end{eqn}
which acts on functions by translating the position by a constant,
\begin{eqn}
U(a) f(x) = f(x-a).
\end{eqn}
We can write 
\begin{eqn}
\partial_a U(a) \Big|_{a \rightarrow 0} = -\partial_x
\end{eqn}
so we say that $\partial_x$ is the \textit{generator} of translations. Up to a constant, we can say that the momentum $p$ is the generator of translation,
\begin{eqn}
p = i \hbar \partial_a U(a) \Big|_{a \rightarrow 0}.
\end{eqn}


\section{Quantum Dynamics}

Temporal evolution will take us from $\ket{\psi(0)} \mapsto \ket{\psi(t)}$. We will consider $t=0$ the present, $t<0$ the past, and $t>0$ the future, and denote the present state $\ket{\psi(0)} = \ket{\psi}$. We will be able to write time evolution as an operator
\begin{eqn}
\ket{\psi} \mapsto \ket{\psi(t)} \equiv U(t) \ket{\psi}.
\end{eqn}
Because the state should be physical and thus normalized at all times, we have
\begin{eqn}
1 = \braket{U(t) \psi}{U(t) \psi} = \bra{\psi} U^\dagger U \ket{\psi}
\end{eqn}
so $U(t)$ is a unitary operator. 

Finally, we can write the Schroedinger equation,
\begin{eqn}
i \hbar \frac{d}{dt} \ket{\psi(t)} = H(t) \ket{\psi(t)}
\end{eqn}
for any physical state $\ket{\psi}$, where $H(t)$ is a Hermitian operator called the Hamiltonian (aka the energy operator). We can rewrite this as
\begin{eqn}
\left[ i \hbar \frac{d}{dt} U(t) \right] \ket{\psi} = \left[ H(t) U(t) \right] \ket{psi},
\end{eqn}
which is true for any $\ket{\psi}$, so we can write the Schroedinger equation instead as an equation for $U(t)$,
\begin{eqn}
i \hbar \frac{d}{dt} U(t) = H(t) U(t), \quad \text{where} \quad U(0) = 1.
\end{eqn}
We can solve this pretty easily -- consider the differential equation
\begin{eqn}
\frac{d}{dt} f(t) = h(t) f(t).
\end{eqn}
We can solve this by rewriting
\begin{eqn}
df / f = h ~ dt
\end{eqn}
which has solution 
\begin{eqn}
\log(f(t) / f(0)) = \int_0^t h(t') dt'.
\end{eqn}
Thus applying this to the equation for $U(t)$ we find the time evolution operator,
\begin{eqn}
U(t) = \exp \left( -\frac{i}{\hbar} \int_0^t H(t') dt' \right). 
\end{eqn}









\end{document}
