% declare document class and geometry
\documentclass[12pt]{article} % use larger type; default would be 10pt
\usepackage[english]{babel} % for hyphenation dictionary
%\setdefaultlanguage{english} % polyglossia command for use with XeTeX / LuaTeX
\usepackage[margin=1in]{geometry} % handle page geometry

% import packages and commands
\input{../header2.tex}

% title information
\title{Phys 221A -- Quantum Mechanics -- Lec12}
\author{UCLA, Fall 2014}
\date{\formatdate{12}{11}{2014}} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed 

\begin{document}
\maketitle


\section{Path integrals and classical mechanics}

Any serious discussion of quantum theory or quantum field theory should first discuss the classical limit. A general path integral can be written
\begin{eqn}
G \sim \int \mathcal{D} \phi \, e^{i S(\phi) / \hbar}
\end{eqn}
where $G$ is a Green's function, $\phi$ is the field (for the propagator just the trajectories), and $S$ is the classical action. We will analyze what this means classically.

Suppose we have a spinless particle of mass $m$ and charge $e$. The physical EM fields are $\v E(\v r, t), \v B(\v r, t)$, and we have the Lorentz force (in cgs units)
\begin{eqn}
\v F_L = e(\v E + \frac{\v v}{c} \times \v B).
\end{eqn}
It turns out that the Lagrangian which determines this force is given by
\begin{eqn}
L(\v r, \vd r, t) = \frac{1}{2} m \vd r^2 - e \phi(\v r, t) + e \frac{\v v}{c} \cdot \v A(\v r, t),
\end{eqn}
where $\phi$ and $\v A$ are the potentials or auxiliary fields,
\begin{eqn}
\v E = - \v \nabla \phi - \frac{1}{c} \vd A, \qquad
\v B = \v \nabla \times \v A.
\end{eqn}
These potentials have a gauge freedom, meaning that different potentials can determine the same physical fields. Specifically, under the transformation
\begin{eqn}
\v A \rightarrow \v A + \v \nabla f, \qquad
\phi \rightarrow \phi - \frac{1}{c} \pd{f}{t},
\end{eqn}
for any arbitrary function $f(\v r, t)$, we will obtain the same physical fields with the same dynamics. We can also verify that this Lagrangian gives us the correct dynamics (the Lorentz force). [See Wikipedia.] 

We can split the Lagrangian into two parts, 
\begin{eqn}
L_0 = \frac{1}{2} m \vd r^2 - e \phi(\v r, t), \qquad
L_g = e \frac{\v v}{c} \cdot \v A(\v r, t).
\end{eqn}
Then the action splits into $S_0$, the ``dynamical action'' and $S_g$, the ``geometric action''. $S_g$ is called the geometric action because if $\v B$ is time-independent then we can choose a gauge where $\v A$ is time-independent, so that 
\begin{eqn}
S_g = \frac{e}{c} \int_i^f \dif{\v r} \cdot \v A(\v r).
\end{eqn}
is a purely geometric quantity. 


\section{Aharonov-Bohm effect}

This is historically interesting because it came a while after Feynman's path integrals, but it comes right out of that so it's interesting that Feynman didn't discover it. 

Suppose we have a static magnetic field $\v B(\v r)$ and a particle constrained to move on a ring (electron in a copper wire, for example). We are interested in how a static $\v B$ field can affect physical properties. In statistical mechanics, we will find that the partition function $Z$ depends on the spectrum $\varepsilon_i$. Also useful for studying heat capacity of the ring $C(\v B)$ and the conductance $g(\v B$). 

If we consider two paths labelled 1 and 2, we can write the propagator
\begin{align}
K &= \int \mathcal{D} [\v r(t)] e^{i S / \hbar} \\
	&\sim K^{(1)} + K^{(2)} \\
	&\sim e^{i S_g^{(1)} / \hbar} \left[ K_0^{(1)} + K_0^{(2)} e^{(i / \hbar) [S_g^{(2)} - S_g^{(1)}]} \right],
\end{align}
so we find that the only really important factor here is the phase difference $\Delta S_g = S_g^{(2)} - S_g^{(1)}$, which is really just the integral around the loop
\begin{eqn}
\Delta S_g = \frac{e}{c} \oint_{2-1} \v A \cdot \dif{\v r}
\end{eqn}
which by Stoke's theorem is just $(e/c) \phi$. 

[missed tiny bit at the end]






\end{document}
