% declare document class and geometry
\documentclass[12pt]{article} % use larger type; default would be 10pt
\usepackage[english]{babel} % for hyphenation dictionary
%\setdefaultlanguage{english} % polyglossia command for use with XeTeX / LuaTeX
\usepackage[margin=1in]{geometry} % handle page geometry

% import packages and commands
\input{../header2.tex}

% title information
\title{Phys 221A -- Quantum Mechanics -- Lec10}
\author{UCLA, Fall 2014}
\date{\formatdate{5}{11}{2014}} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed 

\begin{document}
\maketitle


\section{More on propagators}

Recall that for a time-independent $H$ with eigenstates $\ket{\alpha_i}$ and eigenenergies $E_i$, the propagator is given by
\begin{eqn}
K(\v r, t; \v r', t') = \sum_i \psi_i (\v r) \psi_i^* (\v r') e^{-i E_i (t-t') / \hbar}.
\end{eqn}

\begin{example}[Free particle in 1D]
For example, for a free particle in 1D, we have
\begin{eqn}
\psi_p = \frac{1}{\sqrt{2\pi \hbar}} e^{ipx / \hbar}, \qquad
E_p = \frac{p^2}{2m}
\end{eqn}
which has propagator
\begin{eqn}
K(x, t; x', t') = \int_{-\infty}^\infty \frac{\dif p}{2\pi \hbar} \exp \left[ip (x-x') / \hbar - i \frac{p^2 (t-t')}{2m\hbar} \right].
\end{eqn}
We can evaluate this using the Gaussian integrals
\begin{eqn}
\int_{-\infty}^\infty \dif{x} e^{-x^2} = \sqrt \pi, \qquad \implies \qquad
\int_{-\infty}^\infty \dif{x} e^{-ix^2} = \sqrt{\pi/i} \text{ (or so)},
\end{eqn}
which gives us
\begin{eqn}
K(x,t;x',t') = \sqrt{\frac{m}{2\pi i \hbar (t-t')}} \exp \left[ \frac{im(x-x)^2}{2\hbar (t-t')} \right]
\end{eqn}
for $t \neq t'$. This is just a Gaussian wave packet. Looking at the real part, we have
\begin{eqn}
\Re K \propto \frac{1}{\sqrt{t-t'}} \cos \left[ \frac{m(x-x')^2}{2\hbar(t-t')} - \pi/4 \right].
\end{eqn}
This all tells us that the propagator gets smaller and spreads out over time. 
\end{example}

\begin{example}[1D harmonic oscillator]
For a 1D harmonic oscillator we find a similar propagator,
\begin{eqn}
K(x,t;x',t') = \sqrt{\frac{m\omega}{2\pi i \hbar \sin (\omega(t-t'))}} \exp \left[ i m \omega \frac{(x^2+x'^2) \cos (\omega (t-t')) - 2xx'}{2\hbar \sin (\omega (t-t'))} \right],
\end{eqn}
which you can show reproduces the free propagator when $\omega \rightarrow 0$. This can be used in approximation to model quantum tunneling in various applications. This is called the tight binding approximation. 
\end{example}

Recall that the general propagator is defined by
\begin{eqn}
K(\v r, t; \v r', t') = \matrixel{\v r}{U(t,t')}{\v r'} = \braket{\v r, t}{\v r', t'}
\end{eqn}
where
\begin{eqn}
\ket{\v r, t} \equiv U^\dagger(t) \ket{\v r}.
\end{eqn}
Recall also the composition property
\begin{eqn}
U(t,t'') U(t'',t') = U(t,t'),
\end{eqn}
which implies that
\begin{eqn}
U(t,t') U(t',t) = 1,
\end{eqn}
and thus 
\begin{eqn}
U^\dagger(t,0) = U(0,t). 
\end{eqn}
Thus the Schroedinger equation gives us $\ket{\v r,t}$ at time $t=0$ and $\ket{\v r}$ at time $t$. 

A note on notation. When we write a ket $\ket{\v r}$, $\v r$ is just a vectorial parameter, not an operator. But we will use vectorial operators, so we'll denote that by $\v x$. Then we have
\begin{eqn}
\v x \ket{\v r} = \v r \ket{\v r},
\end{eqn}
or as a wavefunction
\begin{eqn}
\psi_{\v r} (\v R) = \braket{\v R}{\v r} = \delta(\v R - \v r).
\end{eqn}
Of course we can evolve the position operator in time
\begin{eqn}
\v x(t) = U^\dagger(t) \v x U(t)
\end{eqn}
so that we find
\begin{eqn}
\v x(t) \ket{\v r,t} = U^\dagger(t) \v x \underbrace{U(t) U^\dagger(t)}_{=1} \ket{\v r} = \v r \ket{\v r,t}.
\end{eqn}
Furthermore, at fixed time $t$ we have
\begin{eqn}
\braket{\v r, t}{\v r', t} = \delta(\v r - \v r')
\end{eqn}
so that 
\begin{eqn}
1 = \int \dif^3{r} \ket{\v r, t} \bra{\v r, t}
\end{eqn}
for any time $t$, which is just the closure relation. Thus we can think of the propagator as a matrix that takes us from the $t$ basis to the $t'$ basis. 


\section{Feynman path integrals}

If we insert the closure relation into the propagator we have (with $t' < t$) 
\begin{eqn}
K(\v r, t; \v r', t') = \braket{\v r, t}{\v r', t'} = \int \dif^3{r''} \braket{\v r,t}{\v r'',t''} \braket{\v r'',t''}{\v r',t'}.
\end{eqn}
If we take more times $t' < t'' < \dots < t$, we can continue this process and find
\begin{eqn}
K(\v r, t; \v r', t') = \int \dif^3{r''} \cdots \dif^3{r^{(n)}} \braket{\v r,t}{\v r^{(n)},t^{(n)}} \cdots \braket{\v r'', t''}{\v r', t'}.
\end{eqn}

Some history: Feynman and Schwinger both came up with this formalism independently. Amateurs might tell you Feynman is better, but experts know Schwinger is the best. Schwinger was also a professor at UCLA for a long time. He was more idiosyncratic but Feynman was cooler and more intuitive. So you can think of Schwinger as Lennon and Feynman as McCartney. 

We'll continue talking about path integrals next time.





\end{document}
