% declare document class and geometry
\documentclass[12pt]{article} % use larger type; default would be 10pt
\usepackage[margin=1in]{geometry} % handle page geometry

% import packages and commands
\input{../header.tex}


\title{Phys 220A -- Classical Mechanics -- HW01}
\author{UCLA, Fall 2014}
\date{\formatdate{09}{10}{2014}} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed 

\begin{document}
\maketitle


\section*{Problem 1: Math recap (15 pts)}
\begin{description}

\item[(a)]
\textit{
Show that
\begin{equation}
(\v{a} \times \v{b} ) \times \v{c} = (\v{a} \cdot \v{c}) \v{b} - (\v{b} \cdot \v{c}) \v{a}.
\end{equation}
}


\item[(b)]
\textit{
Spherical coordinates $r, \theta, \phi$ are defined as 
\begin{equation}
x = r \sin\theta \cos\phi, \qquad y = r \sin\theta \sin\phi, \qquad z = r \cos\theta.
\end{equation}
Express the cartesian differential operators 
\begin{equation}
\v{\nabla} = (\pd_x, \pd_y, \pd_z), \qquad \v{\nabla}^2 = \pd_x^2 + \pd_y^2 + \pd_z^2
\end{equation}
in spherical coordinates. 
}


\item[(c)]
\textit{
Show the following identities:
\begin{align}
\v{\nabla} \times (\v{\nabla} f(\v{x})) &= 0, \text{ and} \\
\v{\nabla} \cdot (\v{\nabla} \times \v{a}) &= 0.
\end{align}
}


\end{description}



\section*{Problem 2: Conservation is important in a drought! (15 pts)}

\begin{description}

\item[(a)]
\textit{
Show that in an isolated system of $n$ point masses $m_i$ (no external forces), the momentum $\v{p}_\text{tot} = \sum_{i=1}^n \v{p}_i$ is conserved. 
}


\item[(b)]
\textit{
Show that in an isolated system of $n$ point masses $m_i$ (no external forces) the total angular momentum $\v{L}_\text{tot} = \sum_{i=1}^n \v{r}_i \times \v{p}_i$ is conserved. 
}


\item[(c)]
\textit{
What about the total energy of such an isolated system. Is it necessarily conserved? 
}


\end{description}



\section*{Problem 3: Work, work work (15 pts)}

\textit{
Check whether the following forces are conservative by calculating
\begin{equation}
W_{if} = \int_{i \rightarrow f} d\v{x} \cdot \v{F}
\end{equation}
for two different paths. (If you believe the force is non-conservative try to find a path which makes this clear.) 
}

\begin{description}

\item[(a)]
\textit{
For a central force $\v{F} = r^2 \uv{r}$.
}


\item[(b)]
\textit{
For the Lorentz force $\v{F} = -e\v{v} \times \v{B}$ with a constant magnetic field $\v{B}$. 
}


\item[(c)]
\textit{
For a friction force $\v{F} = -b\v{v}$.
}


\end{description}


\section*{Problem 4: Central motion (30 pts)}
\textit{
A planet of mass $m$ is moving around a Sun of mass $M$ subject to Newtonian gravitational force. 
}

\begin{description}

\item[(a)]
\textit{
Write down the Newtonian equation of motion for the planet and the sun, and show that they can be separated into a free motion of the center of mass and a motion in the central potential of a mass point with reduced mass $\mu$.
}


\item[(b)]
\textit{
Show that conservation of angular momentum implies that the motion of the mass point lies in a plane with coordinates $\rho, \phi$.
}


\item[(c)]
\textit{
Use the substitution $u = 1 / \rho$ to write down a differential equation for the trajectory $u = u(\phi)$.
}


\item[(d)]
\textit{
What is the equilibrium solution of this equation? What does it represent?
}


\item[(e)]
\textit{
If the planet is not initially on the equilibrium orbit, there will be small oscillations around the equilibrium point. What is the period of these oscillations?
}


\item[(f)]
\textit{
Assume there is a perturbing potential $V = -B / \rho^2$, calculate the effect of this perturbation on the orbit $u(\phi)$.
}


\end{description}





\end{document}
