% declare document class and geometry
\documentclass[12pt]{article} % use larger type; default would be 10pt
\usepackage[margin=1in]{geometry} % handle page geometry

\input{../header2.tex}

\title{Phys 220A -- Classical Mechanics -- Lec02}
\author{UCLA, Fall 2014}
\date{\formatdate{07}{10}{2014}} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed 

\begin{document}
\setlength{\unitlength}{1mm}
\maketitle


\section{Introduction}

A couple of general comments about Newton's laws. 

\subsection{Something funny about Lorentz force and magnetic fields}

Consider the Lorentz force between two charged particles, given by $\v{F}_{12} = q \v{v}_1 \times \v{B}_2$. Consider particle 1 to lie on the positive $y$ axis with velocity in the positive $y$ direction and particle 2 to lie on the positive $z$ axis with velocity in the positive $x$ axis. Then we find that $\v{F}_{12} \neq 0$ points in the positive $x$ direction yet $\v{F}_{21} = 0$. 

What's going on here? Is this a case of violation of angular momentum conservation? No, in fact the EM field itself carries angular momentum. 


\subsection{Virial Theorem (useful in stat. mech)}

Take some function $f$ and write the time average
\begin{equation}
\avg{f} = \lim_{T \rightarrow \infty} \frac{1}{T} \int_{t_0}^{t_0 + T} f(t) dt.
\end{equation}
Note that there is no dependence on $t_0$ for a bounded quantity,
\begin{equation}
\pd{}{t_0} \avg{f_0} = \lim_{T \rightarrow \infty} \frac{f(t_0 + T) - f(t_0)}{T} = 0
\end{equation}
which vanishes because the numerator in the limit is bounded. 

Next, we hope to obtain an expression for $\avg{T}$ for conservative forces. Notice that 
\begin{equation}
\sum_i m_i \ddot{\v{x}}_i \cdot \v{x} = \sum_i \v{F}_i \cdot \v{x}_i,
\end{equation}
which we can rewrite as 
\begin{equation}
\frac{d}{dt} \left( \sum_i m_i \dot{\v{x}}_i \cdot \v{x} \right) - \underbrace{\sum_i m_i \dot{\v{x}}_i^2}_{2T} = - \sum_i \v{\nabla}_i V \cdot \v{x}_i.
\end{equation}
If the motion is bounded then the derivative goes away when we average, so averaging over time we find
\begin{equation}
2 \avg{T} = \avg{ \sum_i \v{\nabla}_i V \cdot \v{x}_i }.
\end{equation}
This is the Virial theorem. It is especially useful if $V$ is homogeneous, i.e. 
\begin{equation}
V(\lambda \v{x}_1, \dots \lambda \v{x}_n) = \lambda^k V(\v{x}_1, \dots, \v{x}_n)
\end{equation}
for some constant $k$ and any value of $\lambda$. Taking the derivative of this equation w.r.t. $\lambda$ we have
\begin{equation}
\sum_i \v{\nabla}_i V(\lambda \v{x}_1, \dots, \lambda \v{x}_n) \cdot \v{x}_i = k \lambda^{k-1} V(\v{x}_1, \dots, \v{x}_n)
\end{equation}
and setting $\lambda = 1$ we find $\sum_i \v{\nabla} V \cdot \v{x}_i = k V$. Thus the Virial theorem for a homogeneous potential gives us
\begin{equation}
2 \avg{T} = k \avg{V}.
\end{equation}

Let's look at some simple systems. For the harmonic oscillator
\begin{equation}
V = \frac{1}{2} kx^2
\end{equation}
So our parameter from above $k$ is 2, so virial theorem for this system is:
\begin{equation}
\avg{T} = \avg{V}
\end{equation}
For a Keplerian potential it's:
\begin{equation}
V = \frac{\alpha}{r}
\end{equation}
So $k = 1$ and Virial theorem is
\begin{equation}
\avg{T} = -\frac{1}{2} \avg{V}
\end{equation}
\textbf{Note that Virial theorem in general only holds for bound motion with no scattering}


\section{Simple systems}
\subsection{1-D Harmonic Oscillator}
Let's look at the solutions to some systems

Hookes law:
\begin{equation}
F = -kx 
\end{equation}
\begin{equation}
m\ddot{x} = -kx
\end{equation}
\begin{equation}
\ddot{x} + \frac{k}{m} x = 0
\end{equation}
This system has a very general ansatz where $\omega = k / m$
\begin{equation}
x(t) = c_1 \sin\omega t + c_2 \cos\omega t
\end{equation}
\begin{equation}
x(t) = c\sin(\omega t + \phi_0)
\end{equation}
\begin{equation}
c_1 e^{i\omega t} + c_2^* e^{-i \omega t}
\end{equation}
There are 2 integration constants that determine $x$ and $\dot{x}$ at t = 0. We can also write the momentum of this system
\begin{equation}
p = m\dot{x} = mc_1 \omega \cos(\omega t + \phi_0)
\end{equation}
\begin{equation}
p^2 = m^2 \omega^2 (c_1^2 - x^2)
\end{equation}
\begin{equation}
p^2 + \omega x^2 = m^2 \omega^2
\end{equation}
Looking at this in phase space, we get closed circles because this is a conservative field. The origin is the equilibrium point and the circles do not intersect. This changes if there is friction, and for driven harmonic oscillator

\subsection{Particle in Constant Magnetic Field}
Let's set up a constant magnetic field in the $\uv{z}$ direction. Then our forces are 
\begin{equation}
\v{F} = e\v{v} \times \v{B}, \quad \text{or} \quad
F_i = e \epsilon_{ijk} v_j B_k,
\end{equation}
which gives us equations of motion
\begin{eqn}
m\ddot{z} = 0 \quad \implies \quad z = z_0 + v_z t,
\end{eqn}
and
\begin{eqn}
m\ddot{x} = B e \dot{y}, \qquad
m\ddot{y} = -B e\dot{x}.
\end{eqn}
This is a system of two first order differential equations. Defining
\begin{equation}
\omega_B = \frac{eB}{m}, \quad 
S_1 = \dot x, \quad
S_2 = \dot y,
\end{equation}
our equations become
\begin{align}
\dot{S_1} &= \omega_B S_2, \\
\dot{S_2} &= -\omega_B S_1.
\end{align}
Written in vector form, this becomes
\begin{equation}
\d{}{t} \pmat{ S_1 \\ S_2 } = M \pmat{ S_1 \\ S_2 }
\end{equation}
where $M$ is a matrix
\begin{equation}
M = \pmat{ 0 & \omega_B \\ -\omega_B & 0 }
\end{equation}
The solution of course is given by exponentiation
\begin{eqn}
\pmat{ S_1(t) \\ S_2(t) } = R(t) \pmat{ S_1(0) \\ S_2(0) }
\end{eqn}
where
\begin{eqn}
R(t) = \exp (Mt) = 
\begin{pmatrix}
\cos \omega_B t & \sin \omega_B t \\
-\sin \omega_B t & \cos \omega_B t
\end{pmatrix}.
\end{eqn}
Then $x,y$ can be obtained by integration
\begin{align}
x(t) &= \int S_1(t) \dif{t} + x_0, \\
y(t) &= \int S_2(t) \dif{t} + y_0,
\end{align}
which gives us
\begin{align}
x(t) &= \frac{S_1(0)}{\omega_B} \sin \omega_B t - \frac{S_2(0)}{\omega_B} \cos \omega_B t + x_0, \\
y(t) &= \frac{S_1(0)}{\omega_B} \cos \omega_B t + \frac{S_2(0)}{\omega_B} \sin \omega_B t + y_0.
\end{align}
This motion can be described as a drift in the $\uv{z}$ direction with constant velocity and circular motion in the $xy$-plane with radius $R$ and angular frequency $\omega_B = eB/m$, the Larmor frequency. 


\section{Motion in a central potential}

Next we will study motion in a central potential. This is where the potential is a function only of the radial coordinate $r$, most important being the Kepler/Coulomb potential. In the Kepler case it describes the motion of two masses $m_1, m_2$ in their gravitational field. 

The first step is to transform to the center of mass and relative coordinate
\begin{eqn}
\v{X} = \frac{m_1 \v x_1 + m_2 \v x_2}{m_1 + m_2}, \qquad
\v R = \v x_1 - \v x_2.
\end{eqn}
This results in decoupled coordinates: free motion of the center of mass and a central force equation for $R$,
\begin{eqn}
\vdd{X} = 0, \qquad
\mu \vdd{R} = -\v{\nabla} V(R).
\end{eqn}
Some special cases are as follows.


\subsection{Special cases}

\begin{example}[Spherical harmonic oscillator]
$V(r) = \alpha r^2$
\end{example}

\begin{example}[Gravitational/Kepler potential]
$V(r) = - \beta / r$
\end{example}

In general, $V(r)$ can be a complicated function.

\begin{example}[Yukawa potential]
$V(r) = -\frac{\gamma}{r} e^{-r / r_0}$
\end{example}

\begin{example}[Lennard-Jones potential]
$V(r) = \epsilon \left[ (\frac{r_m}{r})^{12} - 2 (\frac{r_m}{r})^6 \right]$, where the first term is a repulsive (Pauli) force, and the second term is an attractive Van der Waals force. 
\end{example}

As we shall see, apart from being simpler than other potentials, the harmonic oscillator and Kepler potentials are special since they allow for closed bounded orbits. 


\subsection{Angular momentum conservation}

In general, angular momentum $\v L = \v x \times \v p$ is conserved for central potentials:
\begin{align}
\vd L &= \vd x \times \v p + \v x \times \vd p \\
	&= \frac{1}{m} \cancel{\v p \times \v p} - \v x \times \v \nabla V \\
	&= 0
\end{align}
since $\v \nabla V \sim \v x$ for central potentials. By choice of coordinates we can choose $\v L = L \uv z$, which gives us motion solely in the $xy$-plane. Note that angular momentum conservation is Kepler's area law---so it's not limited to $1/r$ potentials. 


\subsection{Cylindrical coordinates}

Here we will go to cylindrical coordinates $(x,y,z) \rightarrow (\rho, \phi, z)$, where we have
\begin{eqn}
x = r \cos\phi, \qquad
y = r \sin\phi, \qquad
z = z
\end{eqn}
and velocity
\begin{eqn}
\vd x = \dot r \uv r + r \dot \phi \uv \phi + \cancel{\dot z \uv z}
\end{eqn}
since
\begin{eqn}
\uv r = \frac{1}{r} \pmat{x \\ y}, \qquad
\uv \phi = \frac{1}{r} \pmat{-y \\ x}.
\end{eqn}
So for angular momentum we have
\begin{align}
\v L &= r \uv r \times m (\dot r \uv r + r \dot \phi \uv \phi) \\
	&= m r^2 \dot \phi \uv z
\end{align}
so that $L = m r^2 \dot \phi$ is constant. An application of Kepler's law is that comets have very eccentric orbits and hence spend most of the time in the outer reaches of the solar system. 


\subsection{Integration of Newton's equation}

Note that
\begin{eqn}
\uv{\dot r} = \dot \phi \uv \phi, \qquad
\uv{\dot \phi} = - \dot \phi \uv r,
\end{eqn}
Furthermore, we have the acceleration
\begin{eqn}
\vdd x = (\ddot r - r \dot \phi^2) \uv r + (2 \dot r \dot \phi + r \ddot \phi) \uv \phi + \cancel{\ddot z \uv z},
\end{eqn}
so that our potential equation
\begin{eqn}
m \vdd x = - \v \nabla V(r) = - \pd{V}{r} \uv r
\end{eqn}
can be written as two equations,
\begin{eqn}
m (\ddot r - r \dot \phi^2) = - \pd{V}{r}
\end{eqn}
and
\begin{eqn}
0 = 2 \dot r \dot \phi + r \ddot \phi = \frac{1}{r} \od{}{t} m r^2 \dot \phi.
\end{eqn}
Since $\dot \phi = L^2 / m r^2$ the second equation is just angular momentum conservaion, while for the first equation we obtain a new equation of motion just in $r$,
\begin{eqn}
m \ddot r = -\pd{V}{r} + \frac{L^2}{mr^3}
\end{eqn}
which we can rewrite as
\begin{eqn}
m \ddot r = - \pd{V_\text{eff}}{r}
\end{eqn}
where
\begin{eqn}
V_\text{eff}(r) = V(r) + \frac{L^2}{2mr^2}.
\end{eqn}

Our new equation of motion is 2nd order and nonlinear---would be very difficult to solve explicitly directly from here. But $V(r)$ is a potential for a conservative force, so we have energy conservation
\begin{align}
E &= \frac{1}{2} m \dot r^2 + \frac{1}{2} m r^2 \dot \phi^2 + V(r) \\
	&= \frac{1}{2} m \dot r^2 + V_\text{eff}(r).
\end{align}
One can check that $\od{E}{t} = 0$ using the equation of motion. So instead of solving the 2nd order equation explicitly, we can use energy conservation to reduce the problem to a first order equation which can be solved by integration,
\begin{eqn}
\dot r = \pm \sqrt{ \frac{2}{m} (E - V_\text{eff}(r))},
\end{eqn}
which gives us
\begin{eqn}
\pm (t - t_0) = \int \frac{\dif r}{\sqrt{\frac{2}{m} (E - V_\text{eff}(r))}}.
\end{eqn}
So we have ``reduced to quadrature'', i.e. a single integral. Depending on the form of $V(r)$ this integral may be elementary (readily soluble) or not. Some comments are in order:
\begin{itemize}
\item $\pm (t-t_0) = f(r)$ implies that we need to find $r(t) = f\inv (\pm (t-t_0))$, the functional inverse. 
\item Can also find $\phi(t)$ via integration of $\dot \phi = L / mr^2$,
\begin{eqn}
\phi - \phi_0 = \int \frac{L}{m r(t)^2} \dif{t}.
\end{eqn}
\item The effective potential $V_\text{eff}(r)$ generally comes down from infinity as $1/r^2$ for small $r$ and then goes as $V(r)$ as $r \rightarrow \infty$. For the Kepler potential this gives us a dip beneath the zero which then comes back up and approaches zero as $r \rightarrow \infty$, so that we have unbounded orbits for $E>0$ but bounded orbits for $E < 0$ between $r_\text{min}$ and $r_\text{man}$ where $V_\text{eff}(r_\text{min}) = V_\text{eff}(r_\text{max}) = E$.
\item Instead of finding $r(t)$ we can determine $r(\phi)$ or $\phi(r)$ to obtain the orbit. Using
\begin{eqn}
\od{\phi}{t} = \frac{L}{mr^2}
\end{eqn}
we find
\begin{eqn}
\frac{\dif r}{\sqrt{\frac{2}{m} (E-V_\text{eff})}} = \pm \dif t = \pm \frac{mr^2}{L} \dif \phi,
\end{eqn}
which gives us a solution
\begin{eqn}
\pm (\phi - \phi_0) = \int \dif{r} \frac{L / mr^2}{\sqrt{\frac{2}{m} (E - V_\text{eff})}}.
\end{eqn}
\end{itemize}

\subsection{Bertrand's theorem}

We will state this theorem without proof. It states that the only potentials for which all bounded orbits are closed are the Kepler potential and the radial harmonic oscillator
\begin{eqn}
V_1(r) = - k / r, \qquad
V_2(r) = \frac{1}{2} m \omega^2 r^2.
\end{eqn}







\end{document}
