% declare document class and geometry
\documentclass[12pt]{article} % use larger type; default would be 10pt
\usepackage[margin=1in]{geometry} % handle page geometry

\input{../header2.tex}

\title{Phys 220A -- Classical Mechanics -- Lec10}
\author{UCLA, Fall 2014}
\date{\formatdate{13}{11}{2014}} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed 

\begin{document}
\setlength{\unitlength}{1mm}
\maketitle


\section{Action/Angle Variables}

Consider the Harmonic oscillator
\begin{eqn}
H = \frac{p^2}{2m} + \frac{1}{2} m \omega^2 q^2,
\end{eqn}
where we found that the trajectories in phase space make ellipses centered about the origin. This suggests that we could make a canonical transformation to nice polar coordinates in phase space,
\begin{eqn}
(p,q) \rightarrow (I,\theta), \qquad 
q = \sqrt{\frac{2I}{m\omega}} \sin\theta, \qquad
p = \sqrt{2Im\omega} \cos\theta,
\end{eqn}
This results in a very nice form for the Hamiltonian, $H' = \omega I$, and thus the equations of motion are just
\begin{eqn}
\dot \theta = \omega, \qquad
\dot I = 0.
\end{eqn}
So we can visualize the trajectories as constant motion about a ring on a cylinder, where the ring is at height $I$ and angles are labelled $\theta$. These are called action angle variables.


Finding a canonical transformation that maps in this way we're
essentially looking at bound orbits by definition. So then basically
if I have a canonical transformation from p, q to I, $\theta$ then
these are action angle variables. If we're moving in a potential then
this is a little bit trivial. The point is that I don't know he didn't
finish his scentence. In one dimensions this is not too
exciting but basically the point is that it is not always possible to
do this for all degrees of freedom. This defines a special class of
systems where basically if for a system n degrees of freedom or n
coordinates and n conjugate momenta, if there exists a map where I map
them to I and $\theta$ such that H' is the sum of
%Suppose instead we have system with $N$ degrees of freedom. The system is called \textit{integrable} if there exists a canonical transformation
\begin{eqn}
(q_i, p_i) \rightarrow (I_i, \theta_i), \qquad
H' = \sum_i \omega_i I_i,
\end{eqn}
so that the equations of motion are just
\begin{eqn}
\dot \theta_i = \omega_i, \qquad
\dot I_i = 0.
\end{eqn}
So we have motion in $\theta_i$ on $(S^1)^n = T^n$, the $n$-torus. 

\begin{theorem}[Liouville's theorem]
Suppose we have some canonical coordinates $I_1, \dots, I_N$ with 
\begin{eqn}
\cbr{I_i, H'} = \cbr{I_i, I_j} = 0, \qquad
i,j = 1, \dots, N.
\end{eqn}
Then the system is integrable. 
\end{theorem}



It has N \textbf{integrals of
motion} which are quantities that 
\begin{equation}
\{I_i, H\} = 0
\end{equation}
and which are canonical coordinates. Also that
\begin{equation}
\{I_i, I_j\} = 0 
\end{equation}
Another characterization of an integrable system is that I have N
integrals of motion with all vanishing poisson brackets and that are
linearly independent. So if I have N independent integrals of motion
then the system is integrable. This is also Liouville's theorem. 

So in the central potential problem the we should also have 3 action
angle variables. This also applies to something with infinitely many
degrees of freedom like a field theory. Two more comments before
examples. Integrability is really the exception and not the
rule. They're special. Also, integrability is a global property. What
do I mean by that? If youthink about it, if I have some motion along
the phase space in a non-integrable system what I can do is focus on a small neighborhood of the
path. that will look like a tiny line segment and rotate it so that it
looks like an action angle. In other words, locally with a change of
coordinates I can always make any system into an integrable
system. But only for an infinitesimal amount of time with a
nonintegrable system. With an integrable system every patch maps to an
action angle.

\section{Examples}
\begin{itemize}
\item Harmonic oscillator
\item Motion in central potential
\item Spinning top
\item Toda-lattice - mass points coupled by nearest neighbor
  interactions where my Hamiltonian looks like
\begin{equation}
H = \sum_{i = 1}^{N+1} \frac{1}{2} p_1^2 + \omega^2 e^{q_i - q_{i+1}}
\end{equation}
We can link the last to the first one to make a circle and this is the
SU(N) lattice. This is hard to solve with convential methods because
exponentials are scary. 
\item Cologero-Moses systems
\begin{equation}
H = \sum_i \frac{1}{2} p_i^2 + \sum_{i\ne j}^N V(q_i - q_j)
\end{equation}
\begin{equation}
V_1(q) = \frac{1}{q^2}
\end{equation}
\begin{equation}
V_2(q) = \frac{1}{\sin^2 q}
\end{equation}
\item Study of integrable systems is a full subfield of mathematical
  physics exspecially when applied to field theories
\item The method for finding action angle variables and to check
  whether a system is integrable is Hamilton Jacobi equations
\end{itemize}

\section{Hamilton-Jacobi theory}

Motivations:
\begin{itemize}
\item New formulation of mechanics
\item It gives a practical way to construct action/angle variables (separability) (sometimes)
\item Most efficient way of solving dynamics
\item New way of thinking about the action
\item Geometrical optics / semiclassical approximation (QM)
\end{itemize}

In the Hamilton-Jacobi formulation, we are interested in finding a generating function $F(q,Q,t)$ such that the new Hamiltonian is as simple as possible. What's the simplest Hamiltonian?
\begin{eqn}
K(P,Q,t) \equiv 0.
\end{eqn}
Then $\dot Q = \dot P = 0$ or $Q \equiv \alpha, P \equiv \beta$, so we've just moved the dynamics from the Hamiltonian into the generating function. So we can write, using our results from the generating function formalism,
\begin{eqn}
0 = K(Q_i,P_i,t) = H(p_i,q_i,t) + \pd{F}{t}.
\end{eqn}
This is a canonical transformation of type 1, $F_1 = F$, so we have
\begin{eqn}
p_i = \pd{F}{q_i}, \qquad
P_i = -\pd{F}{Q_i}.
\end{eqn}
Then we can rewrite $K = 0$ as 
\begin{eqn}
H(q_i, \tpd{F(q,Q,t)}{q_i}, t) + \pd{F(q,Q,t)}{t} = 0,
\end{eqn}
which is the \textbf{Hamilton-Jacobi equation} (HJE). As a comparison, recall that Hamilton's equations give us a system of $2N$ first order ODEs. Here in the Hamilton-Jacobi formalism, we have a single first order PDE in $N+1$ variables. 

As a first example, let's look at the free particle $H = p^2 / 2m$. Then the HJE can be written
\begin{eqn}
\frac{1}{2m} \left(\pd{F}{q}\right)^2 + \pd{F}{t} = 0.
\end{eqn}
We see that even for a very simple example, we can end up with a difficult problem because we now have a nonlinear PDE. By inspection, we can write down a solution 
\begin{eqn}
F(q,Q,t) = \frac{m(q-Q)^2}{2t},
\end{eqn}
then we find
\begin{eqn}
p = \pd{F}{q} = \frac{m(q-Q)}{t}, \qquad
P = -\pd{F}{Q} = \frac{m(q-Q)}{t}.
\end{eqn}
Recall that $Q$ and $P$ are supposed to be constant---looking at their forms, we notice that $\alpha \equiv Q = q(0)$ and $\beta \equiv P = p(0)$ and we can write
\begin{eqn}
q(t) = \frac{\beta}{m} t + \alpha.
\end{eqn}
Furthermore, $F$ looks familiar:
\begin{eqn}
F = \frac{p(0)^2}{2m} t,
\end{eqn}
it's just the action! 
\begin{eqn}
S = \int_0^t \frac{1}{2} m \dot q^2 \dif{t} = \frac{\beta^2}{2m} t.
\end{eqn}
So we've found that $F$ evaluated on the solution is the action. 

It turns out that this solution is not unique. Another solution is given by
\begin{eqn}
F = q \sqrt{2m\alpha} - \alpha t,
\end{eqn}
so we have
\begin{eqn}
p = \pd{F}{q} = \sqrt{2m \alpha}, \qquad
P = -\pd{F}{\alpha} = \frac{q\sqrt{2m}}{2\sqrt{\alpha}} - t = \beta,
\end{eqn}
or 
\begin{eqn}
q = \sqrt{\frac{2\alpha}{m}} (t+\beta), \qquad
p = \sqrt{2m\alpha}.
\end{eqn}
[Is this an equivalent action? i.e. under an equivalent Lagrangian?]

We can show that indeed $F = S$. Given the action
\begin{eqn}
S[q(t)] = \int_0^t \dif{t'} \, L(q, \dot q, t'),
\end{eqn}
the variation is just
\begin{eqn}
\delta S = \int_0^t \cancelto{0}{\left( \pd{L}{q} - \od{}{t} \pd{L}{\dot q} \right)} \delta q + \pd{L}{\dot q} \delta q \Big|_0^t 
	= p \Big|_t \delta q \Big|_t
\end{eqn}
so that $\tpd{S}{q|_t} = p |_t$. Furthermore
\begin{eqn}
L = \od{S}{t} = \pd{S}{t} + \pd{S}{q_t} \od{q_t}{t} \big|_t
\end{eqn}
gives us
\begin{eqn}
L(q, \dot q, t) |_t = \pd{S}{t} + p |_t \od{q}{t} \big|_t 
\end{eqn}
so that
\begin{eqn}
\pd{S}{t} = - \left[ p \dot q - L(q, \dot q, t) \right]
\end{eqn}
and thus
\begin{eqn}
\pd{S}{t} = -H(p,q,t) = -H(\pd{S}{q}, q, t),
\end{eqn}
which is just the HJE. 








\end{document}
